The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts

### 2020-05-21 write-up/scifile.tex
update science paper

% In setting up this template for *Science* papers, we've used both
% the \section* command and the \paragraph* command for topical
% divisions.  Which you use will of course depend on the type of paper
% you're writing.  Review Articles tend to have displayed headings, for
% which \section* is more appropriate; Research Articles, when they have
% formal topical divisions at all, tend to signal them with bold text
% that runs into the paragraph, for which \paragraph* is the right
% choice.  Either way, use the asterisk (*) modifier, as shown, to
% suppress numbering.

@@

The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.  As of May 21st, 2020, over 5 million cases have been confirmed worldwide with 1.5 million confirmed cases and over 90 thousand COVID-19 deaths across the United States. This pandemic
has become the focal point of every day life; yet the data landscape for understanding COVID-19 remains limited.  Public databases maintained by Johns Hopkins University~\cite{JHU_Lancet} and the New York Times~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists have attempted to use this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Song2020,Ray2020,2020.IHME}.

@@

Let $N$ denote the population size.  At a fixed time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$. Here, as in survey methodology~\cite{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest. For simplicity, we start by ignoring the dynamic nature of the outbreak and the fact that individuals can recover.  We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).

Here we consider estimates of the number of COVID-19 cases and/or disease prevalence, i.e., the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$. Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.  Then a natural candidate for prevalence is the proportion of positive tests $\bar y = n^{-1} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
Under simple random sampling (SRS), both are unbiased estimators of the  quantities of interest and the variance of the prevalence estimator equals $(N-1)^{-1} \times (1-f)/f \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.

In the current context, the true status is not observed. Instead, individuals are tested for COVID-19 status. Tests are imperfect.  COVID-19 testing is no exception. Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  We suppose disease prevalence is estimated by $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling.

To better understand this issue, we provide the following intuitive and powerful statistical decomposition of the error between $\tilde y_n$ and the true proportion $\bar Y$
\begin{equation}
\label{eq:error}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M},
\end{equation}
where $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ are sampling rates for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.  This extends prior work by \cite{Meng2018} to account for imperfect testing. The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.  The final term in brackets is the \emph{measurement error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence. Figure~2 presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.  We see that the adjustment can be both positive and negative as well as a range of magnitudes.

The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.  Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.  The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.  The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$. We can see here  selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs~\cite{Beesley2020,Beesley2019,Smeden2019}.

Here we highlight two key insights from~\cite{Meng2018} that are relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
$$
\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 \right].
$$
This points to a troubling and paradoxical situation: the error relative to SRS increases as a function of population size.  \cite{Meng2018} termed this the ``Law of Large Populations'' (LLP).  This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $E_{\I} [\rho_{I,Y} ]$ equal) can yield wildly different estimates due to population size.  Large countries like the US may have similar true prevalence rates as smaller countries like the UK.  Even under similar testing strategies, the mean-squared error in the prevalence rate for the US will have substantially more variation in comparison to SRS.  Comparing the US to the UK, for example, the MSE will increase by a factor of almost five.  Thus conclusions drawn from observed case-count records may be not just wrong, but very wrong.

Second, there have been calls for increased testing.  While important, many conflate increased testing capacity with increased quality of testing.  We will discuss this point further in Section \ref{section:discussion}.  As of May 8th, the US has performed $8,412,095$ tests in total.  Figure~1 shows the trajectory of testing per day and positive cases per day in the US.  The US population is roughly $328$ million, meaning the fraction sampled is $f = 0.026$.  After smoothing, the empirical prevalence is $9.1\%$ on May 8th.  Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$.  Under these parameters, the effective sample size is $14$.  Recent proposals~\cite{Siddarth2020} have argued in favor of increasing overall testing capacity.  With more tests available, the relative sampling rate may reduce.  Supposing it drops to $1.2$ then the effective sample size will increase to $299$.  See Appendix~\ref{section:effss} for the mathematical details and calculations in additional scenarios.   The above calculations point to a sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.

\subsection*{Imperfect testing}

We next investigate the impact of measurement error on the effective sample size calculation.  The effective sample size can now be bounded by
$$
n_{eff} \le \frac{f}{1-f} \times \frac{1}{E_{I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
$$
In Section \ref{section:est_dq}, we discuss recent studies that estimate the COVID-19 false positive rate of $0.5$\% and false negative rate $17.2$\%.  Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the effective sample size drops from $14$ to $10$. If the relative sampling fraction drops to $1.2$ then the effective sample size drops from $299$ to $216$.

The above argument shows that the effective sample size is affected in a complex manner by the interaction between selection bias and measurement error.  Here, we discuss the impact of the interaction on the effective sample size when testing capacity is tied to measurement error.  In the current COVID-19 pandemic, there have been well-justified calls for increases in testing capacity.  These increases, however, may come at the cost of increases in false-positive and false-negative rates.  In terms of effective sample size, if the data quality term remains constant, then testing increases may not significantly improve the effective sample size.  Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this increase was associated with  false positive rate increasing from $0.5$\% to $5$\% and the false negative rate rising from $5\%$ to $20$\%.  Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual decrease is by 26\%.  This implies that doubling the sampling rate will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.  In a worst case scenario, if the false negative rate climbs to $0.30$ and the false positive rate to $0.10$ then the factor effective sample size increases by only a factor of $2.3$.

\subsection*{Regrettable rates: complex biases resulting from self-selection}
\label{section:rates}

The data analyst, now wary of estimating prevalence and total counts, pauses and thinks.  They return shortly thereafter, with a follow-up consideration: \emph{Ok, perhaps estimating prevalence is difficult. Certainly, however, we can estimate the rate of growth.  All I want to know is when we hit the point at which the curve flattens and number of deaths decrease.  That can't be too hard, surely!}

Unfortunately, ratio estimators do not cancel errors as now both the numerator and the denominator are uncertain.  This means selection bias and measurement error can have paradoxical effects.  Here we consider the ratio estimators for the relative change in the prevalence rate.  We let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence (across the entire population) at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).  Then the ratio estimator is given by $r = \bar y_t / \bar y_{t-1}$.  We assume both numerator and denominator are the prevalence estimates adjusted for measurement error. If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_t/\bar y_{t-1} = y_t/y_{t-1}$).  We wish to express the error $\frac{\bar y_t}{\bar y_{t-1}} - \frac{\bar Y_{t}}{\bar Y_{t-1}}$ in statistical terms.  Using a Taylor series approximation, we show in the appendix that the error can be expressed approximately as
$$
\begin{aligned}
\frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
&\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
\end{aligned}
$$
where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.  The magnitude of the error depends on the population-level ratio $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate. The second term represents the ``cancellation'' that the naive analyst is hoping will occur.  The analyst's intuition only holds when the difference is equal to zero.  This occurs at times when data quality, sampling fraction, measurement error, and prevalence are all constant across time.

@@

>>>>

@@

The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.  As of May 21st, 2020, over 5 million cases have been confirmed worldwide with 1.5 million cases and over 90 thousand confirmed deaths across the United States. This pandemic
has become the focal point of every day life; yet the data landscape for understanding COVID-19 remains limited.  Public databases maintained by Johns Hopkins University~\cite{JHU_Lancet} and the New York Times~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists have attempted to use this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Song2020,Ray2020,2020.IHME}.

@@

Let $N$ denote the population size.  At a fixed time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$.
Here, for simplicity, we ignore the dynamic nature of the outbreak and the fact that individuals can recover.  As in survey methodology~\cite{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest. Let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$). Suppose we are interested in COVID-19 prevalence, i.e., the population average $\bar Y = Y/N$. Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.  Then a natural candidate for prevalence is the proportion of positive tests $\bar y = n^{-1} \sum_{i=1}^n y_i$.

Tests are imperfect.  COVID-19 testing is no exception which implies a selected individual's true status is not observed. Here we assume the false positive rate (FP) and false negative rate (FN) are known.  Disease prevalence is estimated by $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling (SRS). To understand the impact of selection bias and imperfect testing, we provide a statistical decomposition of the error between $\tilde y_n$ and $\bar Y$
\begin{equation}
\label{eq:error}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M},
\end{equation}
where $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 0)$ are sampling rates for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.  This extends work by \cite{Meng2018} to account for imperfect testing. The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.  The final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence. Figure~2 presents $D_M$ as a function of the relative frequency and odds ratio can be both positive and negative as well as a range of magnitudes~\cite{Beesley2020,Beesley2019,Smeden2019}.

Comparing the mean-squared error (MSE) under a selection mechanism $\I$ with imperfect testing and SRS with perfect testing (i.e.,$FP=FN=0$), we see that
$$
\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right].
$$
This points to a paradoxical situation: the error relative to SRS increases as a function of population size.  This was termed the ``Law of Large Populations'' (LLP)~\cite{Meng2018} and continues to hold in our imperfect testing context.  This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy can yield very different estimates due to population size.  When comparing countries like the US to the UK, the MSE will increase by a factor of almost five.  Thus conclusions drawn from observed case-count records may be not just wrong, but very wrong.

An important follow-up question is ``What is the sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' We show this \emph{effective} sample size can be bounded by
$$
n_{eff} \le \frac{f}{1-f} \times \frac{1}{E_{I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
$$
As of May 8th, the US has performed $8,412,095$ tests.  The US population is roughly $328$ million, so the fraction sampled is $f = 0.026$.  The empirical prevalence is $9.1\%$ on May 8th. A recent study estimates a false positive and negative rate for antibody testing using a lateral flow immunoassay are $0.5$\% and $17.2$\% respectively \cite{Bendavid2020}. Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then using these parameters the effective sample size is $10$.  Recent proposals~\cite{Siddarth2020} have argued for increased testing capacity, which will reduce the relative sampling rate.  Supposing it drops to $1.2$ then the effective sample size will increase to $216$.  This points to a sad state of affairs, saying that the effective sample size even in optimistic scenarios is not better than a small random sample from the population.

Increased testing capacity may also come at the cost of increased in false positive and negative rates.  In terms of effective sample size, if the data quality term remains constant, then testing increases may not significantly improve the effective sample size.  Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this increase was associated with  false positive rate increasing from $0.5$\% to $5$\% and the false negative rate rising from $5\%$ to $20$\%.  Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual decrease is by 26\%.  This implies that doubling the sampling rate will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.  In a worst case scenario, if the false negative rate climbs to $0.30$ and the false positive rate to $0.10$ then the factor effective sample size increases by only a factor of $2.3$.

\paragraph*{Regrettable rates: complex biases resulting from self-selection}

A separate question for communities facing COVID-19 is knowledge of rates of growth in case-count and the number of COVID-19 deaths decrease.  Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence (across the entire population) at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).  Then the ratio estimator is $r = \bar y_t / \bar y_{t-1}$.  Using a Taylor series approximation, the error ${\bar y_t}/{\bar y_{t-1}} - {\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
$$
\begin{aligned}
\frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
&\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
\end{aligned}
$$
where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.  The magnitude of the error depends on the population-level ratio $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate. The second term represents the ``cancellation'' that the naive analyst is hoping will occur.  The analyst's intuition only holds when the difference is equal to zero.  This occurs at times when data quality, sampling fraction, measurement error, and prevalence are all constant across time.

@@

\subsection*{Technical details}

Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.
